\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[french]{babel}
\usepackage{graphicx}
\usepackage{float}
\usepackage{geometry}
\usepackage{ulem}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{multicol}
\geometry{margin=2.5cm}
\begin{multicols}{2}

\begin{document}

UMBB, Faculty of Sciences ING, 1st Year, Semester 1
\\

Department of Computer Science Exam: ISE
\\

Duration : 1h30 Academic Year: 2025–2026
\\

student ID: Name: Group:
\\

Exercise 1 (Course questions:) (5pts)
\\

1. What is the Shell and what is its main function in an operating system?
\\

Solution:
\\

2. What values are stored in the variable \$? and how can these values be interpreted?
\\

Solution:
\\

Exercise 2 (10pts) ⃝
\\

csh
\\

⃝
\\

ssh
\\

Tick the box corresponding to the correct
\\

answer for each question. -0.25 pt for
\\

3. (1 point) Howdoyouredirectthestan-
\\

each incorrect or unanswered ques-
\\

dard output of a command to the end
\\

tion.
\\

of a file named ?
\\

output.txt
\\

⃝
\\

ls > output.txt
\\

1. (1 point) Which command is used to
\\

search for all files in the directory
\\

⃝
\\

ls » output.txt
\\

that were modified dur-
\\

/home/user/
\\

⃝
\\

ls < output.txt
\\

ing the last 7 days?
\\

⃝
\\

ls | output.txt
\\

⃝
\\

find /home/user/ -mtime
\\

7
\\

4. (1 point) How can you execute a com-
\\

mand in the background while sup-
\\

⃝
\\

search /home/user/ -modified
\\

pressing its standard output?
\\

7
\\

⃝
\\

command \&> /dev/null \&
\\

⃝
\\

grep -r "/home/user/"
\\

-mtime 7
\\

⃝
\\

command > /dev/null \&
\\

⃝
\\

locate /home/user/ -modified
\\

⃝
\\

command \&!
\\

7
\\

⃝
\\

command -bg
\\

2. (1 point) Which of the following names
\\

5. (1 point) Which command is used to
\\

represent a shell version?
\\

display running processes and their
\\

⃝
\\

statistics in real time?
\\

sh
\\

⃝ ⃝
\\

ksh top
\\

⃝ ⃝
\\

ps -r head -n 5 report.txt
\\

⃝
\\

stats
\\

⃝
\\

cat -n report.txt | head
\\

-n 5
\\

⃝
\\

monitor
\\

⃝
\\

tail -n 5 report.txt
\\

6. (1 point) How can you change file per-
\\

missions so that the user has all rights,
\\

⃝
\\

nl report.txt | head -n
\\

thegrouphasread-onlyrights, andoth-
\\

5
\\

ers have no rights?
\\

8. (1 point) How can you rename the file
\\

⃝
\\

chmod 744 file.txt
\\

to using the com-
\\

old.txt new.txt
\\

⃝
\\

chmod u=rwx,g=r,o= file.txt
\\

mand line?
\\

⃝
\\

chmod u=rwx,g=,o= file.txt
\\

⃝
\\

rename "old.txt" "new.txt"
\\

⃝
\\

chmod 711 file.txt
\\

⃝
\\

mv "old.txt" "new.txt"
\\

7. (1 point) How can you display the first
\\

⃝
\\

cp "old.txt" "new.txt"
\\

five lines of the file while
\\

report.txt
\\

adding line numbers? ⃝
\\

ren "old.txt" "new.txt"
\\

Exercise 3 (5pts)
\\

Consider a directory named containing several text files with the following struc-
\\

Data
\\

ture: file1.txt: file2.txt:
\\

ID Name Quantity ID Name Quantity
\\

1 Product\_A 10 1 Product\_A 15
\\

2 Product\_B 5 2 Product\_B 10
\\

3 Product\_C 8 3 Product\_C 5
\\

Using the Linux commands , , , , , , and , answer the
\\

cat grep cut sort uniq join split
\\

following questions:
\\

1. Concatenate all the files in the Data directory into a file named TotalResult.txt.
\\

.................................................................................
\\

2. Extract only the lines containing product names.
\\

.................................................................................
\\

3. Sort the products in alphabetical order.
\\

.................................................................................
\\

4. Save the final result in a file named FinalResult.txt.
\\

.................................................................................
\\

Page 2
\\

\subsection*{Solutions}

Exercise 1 – Course Questions (5 pts : 2.5 pts each)

1. What is the Shell and what is its main function in an operating system?
\\

Solution: The shell is a command-line interpreter that acts as an interface
\\

between the user and the operating system. Its main function is to interpret
\\

and execute user commands, manage processes, and provide access to system
\\

services.
\\

Barème: Definition (1 pt) + Role/Function (1.5 pts)
\\

2. What values are stored in the variable \$? and how can these values be interpreted?
\\

Solution: The variable \$? stores the exit status of the last executed command.
\\

A value of 0 indicates successful execution, while any non-zero value indicates
\\

an error or abnormal termination.
\\

Barème: Meaning of \$? (1 pt) + Interpretation (1.5 pts)
\\

Exercise 2 – Multiple Choice Questions (10 pts : 1 pt / question,
\\

-0.25 penalty)
\\

1. Which command is used to search for all files modified during the last 7 days?
\\

√
\\

find /home/user/ -mtime -7
\\

⃝
\\

search /home/user/ -modified 7
\\

⃝
\\

grep -r "/home/user/" -mtime 7
\\

⃝
\\

locate /home/user/ -modified 7
\\

Correct answer: 1 pt
\\

2. Which of the following names represent a shell version?
\\

√
\\

sh
\\

√
\\

ksh
\\

√
\\

csh
\\

⃝
\\

ssh
\\

2 pt correct choice
\\

3. How do you redirect the standard output to the end of a file?
\\

⃝
\\

ls > output.txt
\\

√
\\

ls » output.txt
\\

Page 3
\\

⃝
\\

ls < output.txt
\\

⃝
\\

ls | output.txt
\\

Correct answer: 1 pt
\\

4. Execute a command in the background while suppressing its output?
\\

√
\\

command > /dev/null \&
\\

⃝
\\

command \&> /dev/null
\\

⃝
\\

command \&!
\\

⃝
\\

command -bg
\\

Correct answer: 1 pt
\\

5. Display running processes in real time?
\\

√
\\

top
\\

⃝
\\

ps -r
\\

⃝
\\

stats
\\

⃝
\\

monitor
\\

Correct answer: 1 pt
\\

6. Set permissions: user all rights, group read-only, others none?
\\

√
\\

chmod 740 file.txt
\\

√
\\

chmod u=rwx,g=r,o= file.txt
\\

⃝
\\

chmod 711 file.txt
\\

⃝
\\

chmod 744 file.txt
\\

\subsection*{pt correct answer}

7. Display the first five lines with line numbers?

⃝
\\

head -n 5 report.txt
\\

√
\\

cat -n report.txt | head -n 5
\\

√
\\

nl report.txt | head -n 5
\\

⃝
\\

tail -n 5 report.txt
\\

\subsection*{correct answer}

8. Rename to ?

old.txt new.txt
\\

⃝
\\

rename "old.txt" "new.txt"
\\

√
\\

mv old.txt new.txt
\\

⃝
\\

cp old.txt new.txt
\\

⃝
\\

ren old.txt new.txt
\\

Correct answer: 1 pt
\\

Page 4
\\

Exercise 3 – Linux Commands (5 pts)
\\

1. Concatenate all files into :
\\

TotalResult.txt
\\

cat Data/*.txt > TotalResult.txt
\\

\subsection*{pts}

2. Extract only lines containing product names:

grep "Product\_" TotalResult.txt
\\

1 pt
\\

3. Sort the products alphabetically:
\\

grep "Product\_" TotalResult.txt | sort
\\

1 pt
\\

4. Save the final result in :
\\

FinalResult.txt
\\

grep "Product\_" TotalResult.txt | sort > FinalResult.txt
\\

\subsection*{pts}

Page 5

\end{multicols}

\end{document}